Understanding the Lomb-Scargle Periodogram

### 2017-02-06 PracticalLombScargle.tex
Add postscript

All figures in this paper were produced using Python, and in particular the
Numpy \citep{numpy, numpybook},
Pandas \citep{pandas},
Astropy \citep{Astropy2013},
and Matplotlib \citep{matplotlib} packages.
Code to reproduce these figures is available in Jupyter notebooks at
\url{http://github.com/jakevdp/PracticalLombScargle/}.

@@

\subsubsection{Deconvolution and CLEANing}
With this ability to compute the window function, we might hope to be able to
use it {\it quantitatively} to remove spurious peaks from the observed
power spectrum.
Recall from \eq{g-nonuniform} that the observed data is a point-wise product
of the underlying function and the survey window:
\begin{equation}
  g_{obs}(t) = g(t)W_{\{t_n\}}(t),
\end{equation}
and the convolution theorem tells us that the observed Fourier transform is
a convolution of the true transform and the window transform:
\begin{equation}
  \mathcal{F}\{g_{obs}\} = \mathcal{F}\{g\} \ast \mathcal{F}\{W_{\{t_n\}}\}
\end{equation}
Given this relationship, we might hope to be able to invert this
convolution to recover $\mathcal{F}\{g\}$ directly.
For example, we could write:
\begin{eqnarray}
  g(t) &=& g_{obs}(t) / W_{\{t_n\}}(t)\nonumber\\
  \mathcal{F}\{g\} &=& \mathcal{F}\{g_{obs}/W_{\{t_n\}}\}\nonumber\\
                   &=& \mathcal{F}\{g_{obs}\} \ast \mathcal{F}\{1/W_{\{t_n\}}\}
\end{eqnarray}
Because of the localization of observations, $W_{\{t_n\}}(t)$ is zero for most
values of $t$ and so $1/W_{\{t_n\}}(t)$ and its Fourier transform are not well
defined. For this reason, direct deconvolution is not an option in most
cases of interest---in other words, the deconvolution problem is ill-posed
and does not have a unique solution.

@@

A periodic signal will be reflected in the periodogram by the presence of a
peak of a certain width and height.
In the Fourier view, the precision with which a peak's frequency can be
identified is directly related to the width of the peak; often the
width at half-maximum is used.
This can be formalized more precisely in the least squares interpretation of
the periodogram, in which the inverse of the curvature of the peak is
identified with the uncertainty \citep{ICVG2014}---which in the Bayesian
view amounts to fitting a Gaussian curve to the (exponentiated)
peak \citep{Jaynes1987, Bretthorst88}---and this introduces some first-order
dependence on the number of samples and their signal-to-noise ratio rather
than the peak width alone.

We have motivated from the Fourier window discussions that peak width
scales with the observational baseline; the surprising result is that to
first order the peak width in the periodogram {\it does not depend} on
either the number of observations or their signal-to-noise ratio!
This can be seen visually in \fig{peak-width-height},
which shows periodograms for simulated data with a fixed 4-period baseline,
with varying sample sizes and signal-to-noise ratios.
In all cases, the widths of the primary peak are essentially {\it identical},
regardless of the quality or quantity of data!
Data quality and quantity are reflected in the height of the peak in relation
to the ``background noise'', which speaks to the peak {\it significance} rather
than the precision of frequency estimation.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{fig26_peak_width_height}
  \caption{The effect of the number of points $N$ and the signal-to-noise
    ratio $S/N$ on the expected width and height of the periodogram peak.
    Top panels show the normalized periodogram (\eq{lomb-scargle-normalized}),
    while bottom panels show the PSD-normalized periodogram
    (\eq{lomb-scargle-periodogram})
    Somewhat surprisingly, neither the number of points nor the signal-to-noise
    ratio affects the peak width \figlabel{peak-width-height}}
\end{figure}

Because of this, if you insist on reporting frequency errorbars derived from
peak width, the results can be pretty silly for observations with
long baselines.
For example, going by the peak width in \fig{LINEAR-power}, the
periodogram reveals a period of $2.58014 \pm 0.00004$ hours.
While this is an accurate characterization of the period precision
{\it assuming that peak is the correct one}, it does not capture the
much more relevant uncertainties demonstrated in \sect{failure-modes},
in which we might find ourselves on the wrong peak entirely!
This is why in general, peak widths and Gaussian error bars are not all that
useful for reporting useful or accurate uncertainties in the context of
periodogram analysis.

@@

@@

\subsection{Postscript: Why Lomb-Scargle?}

\begin{itemize}
  \item Advantages: equivalence to least squares, 
  \item Myths: not faster than Schuster, does not correct for irregular sampling, frequencies are not independent, no savings over full model fit at each frequency \citep{ICVG2014}, straight Least Squares treatment not useful \sect{failure-modes}.
\end{itemize}

\todo{wrap up with some witty \& pithy lines of encouragement.}

>>>>

@@

\subsubsection{Deconvolution and CLEANing}
\sectlabel{CLEAN}
With this ability to compute the window function, we might hope to be able to
use it {\it quantitatively} to remove spurious peaks from the observed
power spectrum.
Recall from \eq{g-nonuniform} that the observed data is a point-wise product
of the underlying function and the survey window:
\begin{equation}
  g_{obs}(t) = g(t)W_{\{t_n\}}(t),
\end{equation}
and the convolution theorem tells us that the observed Fourier transform is
a convolution of the true transform and the window transform:
\begin{equation}
  \mathcal{F}\{g_{obs}\} = \mathcal{F}\{g\} \ast \mathcal{F}\{W_{\{t_n\}}\}
\end{equation}
Given this relationship, we might hope to be able to invert this
convolution to recover $\mathcal{F}\{g\}$ directly.
For example, we could write:
\begin{eqnarray}
  g(t) &=& g_{obs}(t) / W_{\{t_n\}}(t)\nonumber\\
  \mathcal{F}\{g\} &=& \mathcal{F}\{g_{obs}/W_{\{t_n\}}\}\nonumber\\
                   &=& \mathcal{F}\{g_{obs}\} \ast \mathcal{F}\{1/W_{\{t_n\}}\}
\end{eqnarray}
Because of the localization of observations, $W_{\{t_n\}}(t)$ is zero for most
values of $t$ and so $1/W_{\{t_n\}}(t)$ and its Fourier transform are not well
defined. For this reason, direct deconvolution is not an option in most
cases of interest---in other words, the deconvolution problem is ill-posed
and does not have a unique solution.

@@

A periodic signal will be reflected in the periodogram by the presence of a
peak of a certain width and height.
In the Fourier view, the precision with which a peak's frequency can be
identified is directly related to the width of the peak; often the
width at half-maximum is used.
This can be formalized more precisely in the least squares interpretation of
the periodogram, in which the inverse of the curvature of the peak is
identified with the uncertainty \citep{ICVG2014}---which in the Bayesian
view amounts to fitting a Gaussian curve to the (exponentiated)
peak \citep{Jaynes87, Bretthorst88}---and this introduces first-order
dependence on the number of samples and their signal-to-noise ratio,
though the uncertainty remains proportional to the peak width.
\todo{explain this difference in terms of FWHM of exponentiated power?}

We have motivated from the Fourier window discussions that peak width is
the inverse of the observational baseline; the surprising result is that to
first order the peak width in the periodogram {\it does not depend} on
either the number of observations or their signal-to-noise ratio!
This can be seen visually in \fig{peak-width-height},
which shows periodograms for simulated data with a fixed 4-period baseline,
with varying sample sizes and signal-to-noise ratios.
In all cases, the widths of the primary peak are essentially {\it identical},
regardless of the quality or quantity of data!
Data quality and quantity are reflected in the height of the peak in relation
to the ``background noise'', which speaks to the peak {\it significance} rather
than the precision of frequency detection.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\textwidth]{fig26_peak_width_height}
  \caption{The effect of the number of points $N$ and the signal-to-noise
    ratio $S/N$ on the expected width and height of the periodogram peak.
    Top panels show the normalized periodogram (\eq{lomb-scargle-normalized}),
    while bottom panels show the PSD-normalized periodogram
    (\eq{lomb-scargle-periodogram})
    Perhaps surprisingly, neither the number of points nor the signal-to-noise
    ratio affects the peak width. \figlabel{peak-width-height}}
\end{figure}

For this reason, if you insist on reporting frequency errorbars derived from
peak width, the results can be pretty silly for observations with
long baselines.
For example, going by the peak width in \fig{LINEAR-power}, the
periodogram reveals a period of $2.58014 \pm 0.00004$ hours, a relative
precision of about 1/1000 percent.
While this is an accurate characterization of the period precision
{\it assuming that peak is the correct one}, it does not capture the
much more relevant uncertainties demonstrated in \sect{failure-modes},
in which we might find ourselves on the wrong peak entirely!
This is why in general, peak widths and Gaussian error bars should generally
be avoided when reporting uncertainties in the context of a periodogram
analysis.

@@

This paper has been a conceptual tour of the Lomb-Scargle periodogram, from
its roots in Fourier analysis, to its equivalence with special cases of
periodic analysis based on least squares model fitting and Bayesian
analysis. I will finish here with a brief summary of the practical
recommendations contained in these pages, along with a somewhat iconoclastic
post-script for further thought.

@@

\subsection{Postscript: Why Lomb-Scargle?}
After considering all of these practical aspects, I think it is worth
revisiting the question of {\it why} astronomers lean toward the
Lomb-Scargle approach rather than the (in many ways simpler) Schuster
periodogram.
In particular, after three decades of use, there seems to be an unrealistic
mythology about the efficacy of the Lomb-Scargle approach among astronomers.
It's common to see statements or implications like the following:
\begin{itemize}
  \item {\it Myth: The Lomb-Scargle periodogram can be computed more efficiently
    than the Schuster periodogram.} Reality: computationally, the two
    are quite similar, and in fact the fastest Lomb-Scargle algorithm currently
    available is based on the Schuster periodogram computed via the
    the NFFT algorithm (see \sect{algorithmic-considerations})
  \item {\it Myth: The Lomb-Scargle periodogram is faster than the least squares
    periodogram because it avoids explicitly solving for model coefficients.}
    Reality: model coefficients can be determined with not much extra
    computation \citep[see discussion in][]{ICVG2014}.
  \item {\it Myth: The Lomb-Scargle periodogram allows analytic computation of
    statistics for periodogram peaks.} Reality: while this is true at
    individual frequencies, it is not true for the more relevant issue of
    peak-value statistics, which must be either approximated or computed by
    bootstrap resampling.
  \item {\it Myth: The Lomb-Scargle periodogram corrects for aliasing due to
    sampling and leads to independent noise at each frequency.}
    Reality: for structured window functions, the Lomb-Scargle
    periodogram has the same partial aliasing issues as the Schuster
    periodogram (see \sect{failure-modes}).
  \item {\it Myth: Bayesian analysis shows that Lomb-Scargle is the optimal
    statistic for detecting periodic signals in data.} Reality: it rather
    shows that Lomb-Scargle is the optimal statistic for
    {\it fitting a sinusoid to data}, which is not the same as saying it
    is optimal for finding a generic, non-sinusoidal signal.
\end{itemize}
With these myths debunked, what advantage remains for Lomb-Scargle?
What advantages would we lose by instead using a Schuster periodogram,
estimating uncertainty, significance, and false alarm probability by
resampling and simulation, as we must for Lomb-Scargle itself?
I think we only lose the advantage of being able to generalize to more
complicated models in the Bayesian sense, in which case you're not really
using Lomb-Scargle at all, but rather a generative Bayesian model for your
data.
The point of correspondance for a sinusoidal model is perhaps useful trivia,
but not a reason to use that model if your data is not known to be sinusoidal.

Conversely, if you use a Schuster approach rather than a Lomb-Scargle approach,
what you gain is the ability to reason quantitatively about the effects of the
survey window function on the resulting periodogram (see \sect{CLEAN}).
While the deconvolution problem is well-known to be ill-posed, there may be a
way forward for cleaned periodograms based on regularized deconvolutions.

In any case, I would contend that there is ample room for practitioners to
question the prevailing folk wisdom of the unspoken advantage of Lomb-Scargle
over approaches based directly on the Fourier transform, such as the Schuster
periodogram.

\subsection{Figures and Code}

All computations and figures in this paper were produced using Python,
and in particular the
Numpy \citep{numpy, numpybook},
Pandas \citep{pandas},
Astropy \citep{Astropy2013},
and Matplotlib \citep{matplotlib} packages.
Code to reproduce these figures is available in Jupyter notebooks at
\url{http://github.com/jakevdp/PracticalLombScargle/}.
